---
title: "Impacts of Growth Media on Lipid Concentrations"
author: "Dr. Sara J. Smith"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: yes
    latex_engine: xelatex
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Load Libraries & Data**

Load necessary libraries and data, clean and refactor data.
```{r, message = F, warning = F}
library(readxl)
library(lmerTest)
library(gamlss)
library(rstatix)
library(DHARMa)
library(performance)
library(see)
library(emmeans)
library(fitdistrplus)
library(PNWColors)
library(tidyverse)
```

```{r, message = F, warning = F}
diam <- read_excel("lipids_myotubeDiam.xlsx") %>%
  mutate(trial = as.factor(replicate),
         day = as.factor(day)) %>%
  select(-replicate)

lipids <- read_excel("lipids_lipids.xlsx") %>%
  mutate(time = fct_relevel(as.factor(time), "MB"))

split(lipids, lipids$lipid) %>% 
  lapply(select, -lipid) %>% 
  list2env(envir = .GlobalEnv)

genes <- read_excel("lipids_geneExpr.xlsx") %>%
  mutate(day = fct_relevel(as.factor(day), "MB"),
         trial = fct_relevel(as.factor(trial), "1"))

split(genes, genes$expr) %>% 
  lapply(select, -expr) %>% 
  list2env(envir = .GlobalEnv)
```

## **Myotube Diameter**

### **Linear Modelling**
```{r}
diam.lmod <- lmer(diameter ~ condition * day + (1|trial), data = diam)
diam.lres <- simulateResiduals(diam.lmod)
plot(diam.lres) # lots of issues again - not surprising 
```

### **Distribution fitting**
```{r}
ggplot(diam, aes(x = diameter)) + 
  geom_histogram() # looks gamma-y

fit.gamma <- fitdist(diam$diameter, "gamma")
fit.lnorm <- fitdist(diam$diameter, "lnorm")
fit.norm <- fitdist(diam$diameter, "norm")
```

```{r}
plot(fit.gamma)
```

```{r}
plot(fit.lnorm)
```

```{r}
plot(fit.norm)
```

```{r}
denscomp(list(fit.gamma, fit.norm, fit.lnorm), legendtext = c("Gamma", "Normal", "Lognormal"))
```

```{r}
gofstat(list(fit.norm, fit.gamma, fit.lnorm))$aic
#lognormal it is
```

### **Lognormal Linear Modelling**
```{r}
diam.mod <- lmer(log(diameter) ~ condition * day + (1|trial), data = diam)
diam.res <- simulateResiduals(diam.mod)
plot(diam.res)
```

```{r}
check_model(diam.mod) # seems fine
```

```{r}
summary(diam.mod) # interaction term is significant
```

Condition: SP and P are sig different
Day: time point are all sig different from day 5
Interaction term: also sig! Need to extract main effects to parse. 

Extracting simple main effects: 
```{r}
diam.condDay.emmeans <- emmeans(diam.mod, pairwise ~ condition | day)
diam.condDay.emmeans
```

```{r}
diam.dayCond.emmeans <- emmeans(diam.mod, pairwise ~ day | condition)
diam.dayCond.emmeans
```


## **Lipids**

### **Linear Modelling**

Model and simulate residuals for each lipid separately
```{r}
for (code in unique(lipids$lipid)) {
  current_data <- get(code)
  
  mod <- lm(concentration ~ condition * time, data = current_data)
  res <- simulateResiduals(fittedModel = mod, plot = FALSE)
  
  assign(paste0(code, ".mod"), mod, envir = .GlobalEnv)
  assign(paste0(code, ".res"), res, envir = .GlobalEnv)
}
```

Plot residuals - no issues unless noted 
```{r}
plot(TAG.res) 
plot(DAG.res)
plot(CHOL.res)
plot(CE.res)
plot(FA.res)
plot(CL.res)
plot(PG.res)
plot(PC.res)
plot(PE.res)
plot(PI.res)
plot(PS.res)
```

### **Summaries & Extracting Estimated Marginal Means or Simple Effects**
```{r}
summary(TAG.mod) 
```

- P and SP are significantly different
- MB and d7 are not sig different 
- interaction between time point and condition is sig

```{r}
tag.emmeans <- emmeans(TAG.mod, pairwise ~ condition | time) 
tag.emmeans
```

```{r}
summary(DAG.mod)
```

- condition, time, and interaction are not sig 

```{r}
summary(CHOL.mod)
```

- time is sig, but not condition or interaction

```{r}
summary(CE.mod)
```

- condition, time, and interaction are sig

```{r}
ce.emmeans <- emmeans(CE.mod, pairwise ~ condition | time)
ce.emmeans
```

```{r}
summary(FA.mod)
```

- time is sig but not condition or interaction

```{r}
summary(FA.mod)
```

- time is sig but not condition or interaction

```{r}
summary(CL.mod)
```

- nothing is sig

```{r}
summary(PG.mod)
```

- condition, time, and interaction are sig

```{r}
pg.emmeans <- emmeans(PG.mod, pairwise ~ condition | time)
pg.emmeans
```

```{r}
summary(PC.mod)
```

- time and condition are sig, but not the interaction

```{r}
pc.noint.mod <- lm(concentration ~ condition + time, data = PC)
summary(pc.noint.mod)
```

```{r}
summary(PE.mod)
```

- time is sig, but not condition or the interaction

```{r}
summary(PI.mod)
```

- nothing is sig

```{r}
summary(PS.mod)
```

- nothing is sig


## **Gene Expression**

### **IDH**
### **Linear Modelling**

```{r}
idh.mod <- lm(deltacq ~ condition + trial, data = idh) # model hits singularity when trial is included as a random effect 
idh.res <- simulateResiduals(idh.mod)
plot(idh.res) # sig issues with quantile deviations 
```

### **Distribution fitting**
```{r}
ggplot(idh, aes(x = deltacq)) + 
  geom_histogram() # v little data so unsure how this will fit 

fit.gamma <- fitdist(idh$deltacq, "gamma")
fit.lnorm <- fitdist(idh$deltacq, "lnorm")
fit.norm <-  fitdist(idh$deltacq, "norm")
```

```{r}
plot(fit.gamma)
```

```{r}
plot(fit.lnorm)
```

```{r}
plot(fit.norm)
```

```{r}
denscomp(list(fit.gamma, fit.norm, fit.lnorm), legendtext = c("Gamma", "Normal", "Lognormal"))
```

```{r}
gofstat(list(fit.norm, fit.gamma, fit.lnorm))$aic
# technically lognormal but I don't like any of these fits 
```

```{r, echo = F, warning = F, message = F}
fits <- fitDist(deltacq, data = idh, type = "realplus")
```
```{r}
fits$fits #BCPE but I doubt a gamlss is a good idea with this little data
```

```{r}
# neither run with a second effect
kruskal.test(deltacq ~ condition, data = idh)

dunn_test(deltacq ~ condition, data = idh)
```

```{r}
idh.gamlss <- gamlss(deltacq ~ condition + trial, data = idh, family = BCPE())

summary(idh.gamlss)

wp(idh.gamlss) # not great

idh.emmeans <- emmeans(idh.gamlss, pairwise ~ condition)
idh.emmeans
```

### **Best approach**

All of the modelling approaches are either inappropriate, NaNs are produced when calculating corrected p-values, or they cannot handle trial as a random effect due to sample size. My suggestion is to go with the approach below: 

```{r}
summary(aov(deltacq ~ condition + trial, data = idh))
```

- condition and trial are NS (same results as non-parametric equivalent, Kruskal Wallis test, above)


### **Myogenin**

### **Linear Modelling**

```{r}
myo.mod <- lmer(deltacq ~ condition * day + (1 | trial), data = myogenin) # almost/near singular 

myo.mod.nore <- lm(deltacq ~ condition * day + trial, data = myogenin) 
myo.res.nore <- simulateResiduals(myo.mod.nore)
plot(myo.res.nore) # quantile deviations sig 

myo.mod.noint <- lmer(deltacq ~ condition + day + (1 | trial), data = myogenin) # still singular 

myo.mod.nore.noint <- lm(deltacq ~ condition + day + trial, data = myogenin) 
myo.res.nore.noint <- simulateResiduals(myo.mod.nore.noint)
plot(myo.res.nore.noint) # no issues 
```

### **Summary & Simple Main Effects**
```{r}
summary(myo.mod.nore.noint)
```

- trial NS

```{r}
myo.mod.noTrial <- lm(deltacq ~ condition * day, data = myogenin) 
myo.res.noTrial <- simulateResiduals(myo.mod.noTrial)
plot(myo.res.noTrial) # sig issues, revert to myo.mod.nore.noint
```

```{r}
myo.emmeans.dayCond <- emmeans(myo.mod.nore.noint, pairwise ~ day | condition)
myo.emmeans.dayCond
```

- within condition, deltacq MB is sig different from day 2, day 5, and day 7 e.g., within SP, MB is sig diff from 2, 5, and 7

```{r}
myo.emmeans.condDay <- emmeans(myo.mod.nore.noint, pairwise ~ condition | day)
myo.emmeans.condDay
```

- across days, condition is NS i.e., at day MB, 2, 5, and 7, deltacq in SP and P are not different from each other 

### **PGC1 Alpha**
### **Linear Modelling**

```{r}
pgc.mod <- lmer(deltacq ~ condition * day + (1 | trial), data = pgc1alpha)
pgc.res <- simulateResiduals(pgc.mod)
plot(pgc.res) # no issues! 
```

### **Summaries & Extracting Simple Effects**

```{r}
summary(pgc.mod)
```

Condition, Day, and Interaction term are sig, so extracting simple main effects 

```{r}
pgc.emmeans.condDay <- emmeans(pgc.mod, pairwise ~ condition | day)
pgc.emmeans.condDay
```

```{r}
pgc.emmeans.dayCond <- emmeans(pgc.mod, pairwise ~ day | condition)
pgc.emmeans.dayCond
```

